Estimating Animal Abundance with N-Mixture Models \\ Using the \emph{R-INLA} Package for \emph{R}

### 2017-05-03 unbranded_manuscript/meehan_et_al_inla_nmix_arxiv.tex
added unbranded manuscript to repo

%% set up %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[article]{jss}
\usepackage{amstext}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{hyperref}

%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author{Timothy D. Meehan\\National\\Audubon Society \And
        Nicole L. Michel\\National\\Audubon Society \And
        H{\aa}vard Rue\\King Abdulla University\\of Science and Technology}
\title{Estimating Animal Abundance with N-Mixture Models Using the \pkg{R-INLA} Package for R}

%% comma-separated
\Plainauthor{Timothy D. Meehan, Nicole L. Michel, Haavard Rue}
%% without formatting
\Plaintitle{Estimating Animal Abundance with N-Mixture Models Using the R-INLA Package for R}
%% a short title
\Shorttitle{\pkg{R-INLA} for N-Mixture Models}

%% an abstract and keywords
\Abstract{
  Successful management of wildlife populations requires accurate estimates of abundance. Abundance estimates can be confounded by imperfect detection during wildlife surveys. N-mixture models enable quantification of detection probability and often produce abundance estimates that are less biased. The purpose of this study was to demonstrate the use of the \pkg{R-INLA} package to analyze N-mixture models and to compare performance of \pkg{R-INLA} to two other common approaches -- \proglang{JAGS} (via the \pkg{runjags} package), which uses Markov chain Monte Carlo and allows Bayesian inference, and \pkg{unmarked}, which uses Maximum Likelihood and allows frequentist inference. We show that \pkg{R-INLA} is an attractive option for analyzing N-mixture models when (1) familiar model syntax and data format (relative to other \proglang{R} packages) are desired, (2) survey level covariates of detection are not essential, (3) fast computing times are necessary (\pkg{R-INLA} is 10 times faster than \pkg{unmarked}, 300 times faster than \proglang{JAGS}), and (4) Bayesian inference is preferred.
}
\Keywords{abundance, detection, \proglang{JAGS}, N-mixture model, \proglang{R}, \pkg{R-INLA}, \pkg{unmarked}, wildlife}
%% without formatting
\Plainkeywords{wildlife, abundance, detection, N-mixture models, R-INLA, R, BUGS}

%% publication information
%% NOTE: This can be left commented and will be filled out by the technical editor
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{2012-06-04}
%% \Acceptdate{2012-06-04}

%% The address of (at least) one author should be given
\Address{
  Timothy D. Meehan\\
  National Audubon Society\\
  Boulder, Colorado USA\\
  E-mail: \email{tmeehan@audubon.org}\\
\\
  Nicole L. Michel\\
  National Audubon Society\\
  San Francisco, California USA\\
  E-mail: \email{nmichel@audubon.org}\\
  \\
  H{\aa}vard Rue\\
  King Abdulla University
  of Science and Technology\\
  Thuwal, Saudi Arabia\\
  E-mail: \email{haavard.rue@kaust.edu.sa }
}
%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% start main document %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

%% introduction
\section[Introduction]{Introduction}
\subsection[Background]{Background}
Successful management of wildlife species requires accurate estimates of abundance \citep{Yoccoz_Nichols_Boulinier_2001}. One common method for estimating animal abundance is direct counts \citep{Pollock_Nichols_Simons_Farnsworth_Bailey_Sauer_2002}. Efforts to obtain accurate abundance estimates via direct counts can be hindered by the cryptic nature of many wildlife species and by other factors such as observer expertise, weather, and habitat structure \citep{Denes_Silveira_Beissinger_2015}. The lack of perfect detection in wildlife surveys is common, and can cause abundance to be greatly underestimated \citep{Wenger_Freeman_2008, Joseph_Elkin_Martin_Possingham_2009}.

@@

The development of metapopulation designs and N-mixture models represents a significant advance in quantitative wildlife ecology. However, there are practical issues that sometimes act as barriers to adoption. Many of the examples of N-mixture models in the wildlife literature have employed Bayesian modeling software such as \proglang{WinBUGS}, \proglang{OpenBUGS}, or \proglang{JAGS} \citep{plummer2003jags,Lunn_Jackson_Best_Thomas_Spiegelhalter_2012}. These are extremely powerful and flexible platforms for analyzing hierarchical models, but they come with a few important challenges.  First, many wildlife biologists are not accustomed to coding statistical models using the \proglang{BUGS} modeling syntax. While there are several outstanding resources aimed at teaching this skill \citep{Royle_Dorazio_2008, Kery_2010, Kery_Schaub_2011, Kery_Royle_2015} it is, nonetheless, a considerable commitment. Second, while Markov chain Monte Carlo (MCMC) chains converge quickly for relatively simple N-mixture models, convergence for more complex models can take hours to days, or may not occur at all.

There are other tools available for analyzing N-mixture models that alleviate some of these practical issues.  The \pkg{unmarked} package \citep{Fiske_Chandler_others_2011} for \proglang{R} statistical computing software \citep{R_Core_Team_2016} offers several options for analyzing N-mixture models within a frequentist framework, with the capacity to accommodate overdispersion and dynamic populations. The model coding syntax used in \pkg{unmarked} is a simple extension of the standard \proglang{R} modeling syntax. Models are analyzed using Maximum Likelihood (ML), so model analysis is often completed in a fraction of the time taken using an MCMC approach. The familiar model syntax and rapid model evaluation of \pkg{unmarked} has undoubtedly contributed to the broader adoption of N-mixture models by wildlife biologists. However, it comes at a cost -- loss of the intuitive inferential framework associated with Bayesian analysis.

Here we discuss analysis of N-mixture models using the \pkg{R-INLA} package \citep{Martins_Simpson_Lindgren_Rue_2013,Rue_Riebler_Sorbye_Illian_Simpson_Lindgren_2017} for \proglang{R}. The \pkg{R-INLA} package uses integrated nested Laplace approximation (INLA) to derive posterior distributions for a large class of Bayesian statistical models that can be formulated as latent Gaussian models \citep{Rue_Martino_Chopin_2009, Lindgren_Rue_Lindstrom_2011}. INLA was developed to allow estimation of posterior distributions in a fraction of the time taken by MCMC. Like \pkg{unmarked}, the model syntax used in the \pkg{R-INLA} package is a straightforward extension of the modeling syntax commonly used in \proglang{R}. Also, like \pkg{unmarked}, the computational cost of analyzing models with \pkg{R-INLA} is relatively low. The \pkg{R-INLA} approach is different from \pkg{unmarked} in that inference about model parameters falls within a Bayesian framework.

\subsection[Overall objectives]{Overall objectives}
The purpose of this manuscript is to demonstrate analysis of N-mixture models using the \pkg{R-INLA} package. In the process, we employ simulated and real count datasets, and analyze them using \proglang{JAGS}, via the \pkg{runjags} package \citep{Denwood_2016} for \proglang{R}, the \pkg{unmarked} package, and the \pkg{R-INLA} package. In each case, we demonstrate how data are formatted, how models are specified, how model estimates compare to simulation inputs, and how methods compare in terms of computational performance. We also explore a limitation of the \pkg{R-INLA} approach related to model specification. Specifically, while it is possible to specify survey level covariates for detection using \proglang{JAGS} and \pkg{unmarked}, this is not possible using \pkg{R-INLA}. Rather, survey level covariates of detection must be averaged to the site or site-year level. Using an averaged detection covariate does allow accounting for site-level differences in survey conditions, should they occur. However, in the process of averaging, information related to detection within a site or site-year combination is discarded, which could lead to biased detection and abundance estimates under certain conditions.

Much of the code used to conduct analyses is shown in the body of this manuscript. However, some repeated code, and code related to generating tables and figures, is not shown for brevity. All code can be accessed via \url{https://github.com/tmeeha/inlaNMix}. See \url{https://r-inla.org} to connect with the community around the development of \pkg{R-INLA} and its application to geostatistics, biostatistics, epidemiology, and econometrics \citep{Lindgren_Rue_2015,Blangiardo_Cameletti_2015}.

@@

\begin{CodeInput}
R> data4sim <- function(n.sites = 72,   # number study sites
+                       n.surveys = 3,  # number replicate surveys
+                       n.years = 9,    # number years
+                       lam.b0 = 2.0,   # intercept for log lambda
+                       lam.b1 = 2.0,   # slope for log lambda cov 1
+                       lam.b2 = -3.0,  # slope for log lambda cov 2
+                       lam.yr = 1.0,   # slope for log lambda year
+                       p.b0 = 1.0,     # intercept for logit p
+                       p.b1 = -2.0,    # slope for logit p cov 1
+                       p.b3 = 1.0,     # slope for logit p cov 3
+                       disp.size = 3.0 # size of overdispersion
+                       ){
+  # setup
+  if(n.years %% 2 == 0) {n.years <- n.years + 1}     # make years odd
+  N.tr <- array(dim = c(n.sites, n.years))           # for abund
+  y1 <- array(dim = c(n.sites, n.surveys, n.years))  # for ex1 counts
+  y2 <- array(dim = c(n.sites, n.surveys, n.years))  # for ex2 counts
+  # abundance covariate values
+  x.lam.1 <- array(as.numeric(scale(runif(n = n.sites, -0.5, 0.5), 
+                   scale = F)), dim = c(n.sites, n.years))
+  x.lam.2 <- array(as.numeric(scale(runif(n = n.sites, -0.5, 0.5), 
+                   scale = F)), dim = c(n.sites, n.years))
+  yrs <- 1:n.years
+  yrs <- (yrs - mean(yrs)) / (max(yrs - mean(yrs))) / 2
+  yr <- array(rep(yrs, each = n.sites), dim = c(n.sites, n.years))
+  # fill abundance array
+  lam.tr <- exp(lam.b0 + lam.b1*x.lam.1 + lam.b2*x.lam.2 + lam.yr*yr)
+  for(i in 1:n.sites){
+    for(k in 1:n.years){
+    N.tr[i, k] <- rnbinom(n = 1, mu = lam.tr[i, k], size = disp.size)
+  }}
+  # detection covariate values
+  x.p.1 <- array(x.lam.1[,1], dim = c(n.sites, n.surveys, n.years))
+  x.p.3 <- array(as.numeric(scale(runif(n = n.sites*n.surveys*n.years, 
+                                        -0.5, 0.5), scale=F)), 
+                 dim=c(n.sites, n.surveys, n.years))
+  # average x.p.3 per site-year
+  x.p.3.mean <- apply(x.p.3, c(1,3), mean, na.rm=F)
+  out1 <- c()
+  for(k in 1:n.years){
+   chunk1 <- x.p.3.mean[,k]
+   chunk2 <- rep(chunk1, n.surveys)
+   out1 <- c(out1, chunk2)
+  }
+  x.p.3.arr <- array(out1, dim=c(n.sites, n.surveys, n.years))
+  # fill count array with site-yr x.p.3
+  p.tr1 <- plogis(p.b0 + p.b1*x.p.1 + p.b3*x.p.3.arr)
+  for (i in 1:n.sites){
+    for (k in 1:n.years){
+      for (j in 1:n.surveys){
+        y1[i,j,k] <- rbinom(1, size = N.tr[i,k], prob = p.tr1[i,j,k])
+  }}}
+  # fill count array with site-surv-yr x.p.3
+  p.tr2 <- plogis(p.b0 + p.b1 * x.p.1 + p.b3 * x.p.3)
+  for (i in 1:n.sites){
+    for (k in 1:n.years){
+      for (j in 1:n.surveys){
+        y2[i,j,k] <- rbinom(1, size = N.tr[i,k], prob = p.tr2[i,j,k])
+  }}}
+  # return data
+  return(list(n.sites = n.sites, n.surveys = n.surveys, n.years = n.years,
+              x.p.1 = x.p.1[,1,1], x.p.3 = x.p.3, x.p.3.mean = x.p.3.mean,
+              x.p.3.arr = x.p.3.arr, x.lam.1 = x.lam.1[,1],
+              x.lam.2 = x.lam.2[,1], yr = yr[1,], y1 = y1, y2 = y2,
+              lam.tr = lam.tr, N.tr = N.tr))
+  } #end function
R> set.seed(12345) # make results reproducible
R> sim.data <- data4sim() # create datasets
\end{CodeInput}

\subsection[Real data]{Real data}
In addition to simulated data, we also demonstrate the use of \pkg{R-INLA} with a real dataset in Example III. This dataset comes from a study by \cite{Kery_Royle_Schmid_2005} and is publicly available as part of the \pkg{unmarked} package. The dataset includes mallard duck (\emph{Anas platyrhynchos}) counts, conducted at 239 sites on 2 or 3 occasions during the summer of 2002 as part of a Swiss program that monitors breeding bird abundance (Monitoring H\"{a}ufige Brutv\"{o}gel). In addition to counts, the dataset also includes 2 site-survey covariates related to detection (survey effort and survey date), and 3 site level covariates related to abundance (route length, elevation, and forest cover). Full dataset details are given in \cite{Kery_Royle_Schmid_2005}.

%% example 1
\section[Example I]{Example I}
\subsection[Goals]{Goals}
In Example I, we demonstrate the use of \pkg{R-INLA} and compare use and performance to similar analyses using \proglang{JAGS} and \pkg{unmarked}. In this exercise, the functional forms of \proglang{JAGS}, \pkg{unmarked}, and \pkg{R-INLA} models match the data generating process.  Specifically, we used the covariate x.p.3.mean to generate the count matrix y1 and analyzed the data with models that use x.p.3.mean as a covariate. This exercise was intended to demonstrate the differences and similarities in use, computation time, and estimation results across the three methods when the specified model was reasonably close to the data generating process.

\subsection[Analysis with JAGS]{Analysis with \proglang{JAGS}}
We first analyzed the simulated data using \proglang{JAGS}, via the \pkg{runjags} package. In defining the model, we specified a Negative Binomial distribution for the abundance component, and used vague normal priors for the intercepts and the global effects of the covariates of $\lambda$ and $p$.

\begin{CodeInput}
R> jags.model.string <- "
+    model {
+    # priors
+    intP ~ dnorm(0, 0.01)       # detection intercept
+    bCov1P ~ dnorm(0, 0.01)     # detection cov 1 effect
+    bCov3P ~ dnorm(0, 0.01)     # detection cov 3 effect
+    intLam ~ dnorm(0, 0.01)     # lambda intercept
+    bCov1Lam ~ dnorm(0, 0.01)   # lambda cov 1 effect
+    bCov2Lam ~ dnorm(0, 0.01)   # lambda cov 2 effect
+    bYr ~ dnorm(0, 0.01)        # year effect
+    overDisEst ~ dunif(0, 5)    # overdispersion size
+    # abundance component
+    for (k in 1:nYears){
+      for (i in 1:nSites){
+        N[i, k] ~ dnegbin(prob[i, k], overDisEst) # negative binomial
+        prob[i, k] <- overDisEst / (overDisEst + lambda[i, k])
+        log(lambda[i, k]) <- intLam + (bCov1Lam * x.lam.1[i]) + 
+                             (bCov2Lam * x.lam.2[i]) + (bYr * yr[k])
+    # detection component
+        for (j in 1:nSurveys){
+          y[i, j, k] ~ dbin(p[i,j,k], N[i,k])
+          p[i, j, k] <- exp(lp[i,j,k]) / (1 + exp(lp[i,j,k]))
+          lp[i, j, k] <- intP + (bCov1P * x.p.1[i]) + (bCov3P * x.p.3[i, k])
+        } # close j loop
+      } # close i loop
+    } # close k loop
+  } # close model loop
+  "
\end{CodeInput}

Next, we defined the parameters to be monitored during the MCMC runs, bundled the data for \proglang{JAGS}, and created a function for drawing random initial values for the model parameters. The initial values for abundance were made to avoid values of "NA" and zero, as these would cause computational problems.

\begin{CodeInput}
R> params <- c("intP", "bCov1P", "bCov3P", "intLam", "bCov1Lam","bCov2Lam",
+              "bYr", "overDisEst") # parameters to monitor
R> jags.data <- list(y = sim.data$y1, x.lam.1 = sim.data$x.lam.1,
+               x.lam.2 = sim.data$x.lam.2, yr = sim.data$yr,
+               x.p.1 = sim.data$x.p.1, x.p.3 = sim.data$x.p.3.mean,
+               nSites = sim.data$n.sites, nSurveys = sim.data$n.surveys,
+               nYears = sim.data$n.years)
R> N.init <- sim.data$y1 # initial count values
R> N.init[is.na(N.init)] <- 1 # clean up NA's
R> N.init <- apply(N.init, c(1, 3), max) + 1 # zero values cause trouble
R> inits <- function() list(N = N.init, intLam = rnorm(1, 0, 0.01),
+               intP = rnorm(1, 0, 0.01), bCov1P = rnorm(1, 0, 0.01),
+               bCov2Lam = rnorm(1, 0, 0.01), bCov1Lam = rnorm(1, 0, 0.01),
+               bCov3P = rnorm(1, 0, 0.01), bYr = rnorm(1, 0, 0.01),
+               overDisEst = runif(1, 0.5, 2.5))
\end{CodeInput}

Finally, we set the run parameters and started the MCMC process. Run parameters were chosen such that MCMC diagnostics indicated converged chains (potential scale reduction factors < 1.05) and reasonably robust posterior distributions (effective sample sizes > 1000). Note that the recommended number of effective samples for particularly robust inference is closer to 6000 \citep{Gong_Flegal_2016}. So MCMC processing times reported here could be considered low estimates.

\begin{CodeInput}
R> nc <- 3; na <- 2500; nb <- 2500; ni <- 5000; nt <- 10 # run parameters
R> out.jags <- run.jags(model = jags.model.string, data = jags.data,
+                       monitor = params, n.chains = nc, inits = inits,
+                       burnin = nb, adapt = na, sample = ni, thin = nt,
+                       modules = "glm on", method = "parallel")
\end{CodeInput}

Mean parameter estimates from the \proglang{JAGS} model were reasonably close to, and not significantly different from, the input values used to generate the data (Fig. \ref{fig:fig1}). The potential scale reduction factor for all variables was < 1.01, and the effective sample size for all variables was > 2409. The simulation ran in parallel on 3 virtual cores, 1 MCMC chain per core, and took approximately 2104 seconds.

>>>>

%% set up %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass{article}
\usepackage{amstext}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{natbib}
\bibliographystyle{plainnat}
\usepackage{authblk}
\usepackage[margin=28mm]{geometry}
\usepackage{titling}
\newcommand{\subtitle}[1]{%
	\posttitle{%
		\par\end{center}
	\begin{center}\large#1\end{center}
	\vskip4mm}%
}
\title{	Estimating Animal Abundance with N-Mixture Models \\ Using the \emph{R-INLA} Package for \emph{R}}

\subtitle{DRAFT: NOT YET PEER REVIEWED}
\author{Timothy D. Meehan$^1$, Nicole L. Michel$^2$, H{\aa}vard Rue$^3$}
\date{
	$^1$\emph{National Audubon Society, Boulder, CO USA\\}
	$^2$\emph{National Audubon Society, San Francisco, CA USA\\}
	$^3$\emph{King Abdulla University of Science and Technology, Thuwal, Saudi Arabia\\}
}

% start   %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
\maketitle

\begin{abstract}
	\fontsize{10pt}{11pt}\selectfont
	Successful management of wildlife populations requires accurate estimates of abundance. Abundance estimates can be confounded by imperfect detection during wildlife surveys. N-mixture models enable quantification of detection probability and often produce abundance estimates that are less biased. The purpose of this study was to demonstrate the use of the \emph{R-INLA} package to analyze N-mixture models and to compare performance of \emph{R-INLA} to two other common approaches -- \emph{JAGS} (via the \emph{runjags} package), which uses Markov chain Monte Carlo and allows Bayesian inference, and \emph{unmarked}, which uses Maximum Likelihood and allows frequentist inference. We show that \emph{R-INLA} is an attractive option for analyzing N-mixture models when (1) familiar model syntax and data format (relative to other \emph{R} packages) are desired, (2) survey level covariates of detection are not essential, (3) fast computing times are necessary (\emph{R-INLA} is 10 times faster than \emph{unmarked}, 300 times faster than \emph{JAGS}), and (4) Bayesian inference is preferred.
\end{abstract}

%% introduction
\fontsize{11pt}{12pt}\selectfont
\section[Introduction]{Introduction}
\subsection[Background]{Background}
Successful management of wildlife species requires accurate estimates of abundance \citep{Yoccoz_Nichols_Boulinier_2001}. One common method for estimating animal abundance is direct counts \citep{Pollock_Nichols_Simons_Farnsworth_Bailey_Sauer_2002}. Efforts to obtain accurate abundance estimates via direct counts can be hindered by the cryptic nature of many wildlife species and by other factors such as observer expertise, weather, and habitat structure \citep{Denes_Silveira_Beissinger_2015}. The lack of perfect detection in wildlife surveys is common, and can cause abundance to be greatly underestimated \citep{Wenger_Freeman_2008, Joseph_Elkin_Martin_Possingham_2009}.

@@

The development of metapopulation designs and N-mixture models represents a significant advance in quantitative wildlife ecology. However, there are practical issues that sometimes act as barriers to adoption. Many of the examples of N-mixture models in the wildlife literature have employed Bayesian modeling software such as \emph{WinBUGS}, \emph{OpenBUGS}, or \emph{JAGS} \citep{plummer2003jags,Lunn_Jackson_Best_Thomas_Spiegelhalter_2012}. These are extremely powerful and flexible platforms for analyzing hierarchical models, but they come with a few important challenges.  First, many wildlife biologists are not accustomed to coding statistical models using the \emph{BUGS} modeling syntax. While there are several outstanding resources aimed at teaching this skill \citep{Royle_Dorazio_2008, Kery_2010, Kery_Schaub_2011, Kery_Royle_2015} it is, nonetheless, a considerable commitment. Second, while Markov chain Monte Carlo (MCMC) chains converge quickly for relatively simple N-mixture models, convergence for more complex models can take hours to days, or may not occur at all.

There are other tools available for analyzing N-mixture models that alleviate some of these practical issues.  The \emph{unmarked} package \citep{Fiske_Chandler_others_2011} for \emph{R} statistical computing software \citep{R_Core_Team_2016} offers several options for analyzing N-mixture models within a frequentist framework, with the capacity to accommodate overdispersion and dynamic populations. The model coding syntax used in \emph{unmarked} is a simple extension of the standard \emph{R} modeling syntax. Models are analyzed using Maximum Likelihood (ML), so model analysis is often completed in a fraction of the time taken using an MCMC approach. The familiar model syntax and rapid model evaluation of \emph{unmarked} has undoubtedly contributed to the broader adoption of N-mixture models by wildlife biologists. However, it comes at a cost -- loss of the intuitive inferential framework associated with Bayesian analysis.

Here we discuss analysis of N-mixture models using the \emph{R-INLA} package \citep{Martins_Simpson_Lindgren_Rue_2013,Rue_Riebler_Sorbye_Illian_Simpson_Lindgren_2017} for \emph{R}. The \emph{R-INLA} package uses integrated nested Laplace approximation (INLA) to derive posterior distributions for a large class of Bayesian statistical models that can be formulated as latent Gaussian models \citep{Rue_Martino_Chopin_2009, Lindgren_Rue_Lindstrom_2011}. INLA was developed to allow estimation of posterior distributions in a fraction of the time taken by MCMC. Like \emph{unmarked}, the model syntax used in the \emph{R-INLA} package is a straightforward extension of the modeling syntax commonly used in \emph{R}. Also, like \emph{unmarked}, the computational cost of analyzing models with \emph{R-INLA} is relatively low. The \emph{R-INLA} approach is different from \emph{unmarked} in that inference about model parameters falls within a Bayesian framework.

\subsection[Overall objectives]{Overall objectives}
The purpose of this manuscript is to demonstrate analysis of N-mixture models using the \emph{R-INLA} package. In the process, we employ simulated and real count datasets, and analyze them using \emph{JAGS}, via the \emph{runjags} package \citep{Denwood_2016} for \emph{R}, the \emph{unmarked} package, and the \emph{R-INLA} package. In each case, we demonstrate how data are formatted, how models are specified, how model estimates compare to simulation inputs, and how methods compare in terms of computational performance. We also explore a limitation of the \emph{R-INLA} approach related to model specification. Specifically, while it is possible to specify survey level covariates for detection using \emph{JAGS} and \emph{unmarked}, this is not possible using \emph{R-INLA}. Rather, survey level covariates of detection must be averaged to the site or site-year level. Using an averaged detection covariate does allow accounting for site-level differences in survey conditions, should they occur. However, in the process of averaging, information related to detection within a site or site-year combination is discarded, which could lead to biased detection and abundance estimates under certain conditions.

Much of the code used to conduct analyses is shown in the body of this manuscript. However, some repeated code, and code related to generating tables and figures, is not shown for brevity. All code can be accessed via https://github.com/tmeeha/inlaNMix. See https://r-inla.org to connect with the community around the development of \emph{R-INLA} and its application to geostatistics, biostatistics, epidemiology, and econometrics \citep{Lindgren_Rue_2015,Blangiardo_Cameletti_2015}.

@@

\begin{verbatim}
R> data4sim <- function(n.sites = 72,   # number study sites
+                       n.surveys = 3,  # number replicate surveys
+                       n.years = 9,    # number years
+                       lam.b0 = 2.0,   # intercept for log lambda
+                       lam.b1 = 2.0,   # slope for log lambda cov 1
+                       lam.b2 = -3.0,  # slope for log lambda cov 2
+                       lam.yr = 1.0,   # slope for log lambda year
+                       p.b0 = 1.0,     # intercept for logit p
+                       p.b1 = -2.0,    # slope for logit p cov 1
+                       p.b3 = 1.0,     # slope for logit p cov 3
+                       disp.size = 3.0 # size of overdispersion
+                       ){
+  # setup
+  if(n.years %% 2 == 0) {n.years <- n.years + 1}     # make years odd
+  N.tr <- array(dim = c(n.sites, n.years))           # for abund
+  y1 <- array(dim = c(n.sites, n.surveys, n.years))  # for ex1 counts
+  y2 <- array(dim = c(n.sites, n.surveys, n.years))  # for ex2 counts
+  # abundance covariate values
+  x.lam.1 <- array(as.numeric(scale(runif(n = n.sites, -0.5, 0.5), 
+                   scale = F)), dim = c(n.sites, n.years))
+  x.lam.2 <- array(as.numeric(scale(runif(n = n.sites, -0.5, 0.5), 
+                   scale = F)), dim = c(n.sites, n.years))
+  yrs <- 1:n.years
+  yrs <- (yrs - mean(yrs)) / (max(yrs - mean(yrs))) / 2
+  yr <- array(rep(yrs, each = n.sites), dim = c(n.sites, n.years))
+  # fill abundance array
+  lam.tr <- exp(lam.b0 + lam.b1*x.lam.1 + lam.b2*x.lam.2 + lam.yr*yr)
+  for(i in 1:n.sites){
+    for(k in 1:n.years){
+    N.tr[i, k] <- rnbinom(n = 1, mu = lam.tr[i, k], size = disp.size)
+  }}
+  # detection covariate values
+  x.p.1 <- array(x.lam.1[,1], dim = c(n.sites, n.surveys, n.years))
+  x.p.3 <- array(as.numeric(scale(runif(n = n.sites*n.surveys*n.years, 
+                                        -0.5, 0.5), scale=F)), 
+                 dim=c(n.sites, n.surveys, n.years))
+  # average x.p.3 per site-year
+  x.p.3.mean <- apply(x.p.3, c(1,3), mean, na.rm=F)
+  out1 <- c()
+  for(k in 1:n.years){
+   chunk1 <- x.p.3.mean[,k]
+   chunk2 <- rep(chunk1, n.surveys)
+   out1 <- c(out1, chunk2)
+  }
+  x.p.3.arr <- array(out1, dim=c(n.sites, n.surveys, n.years))
+  # fill count array with site-yr x.p.3
+  p.tr1 <- plogis(p.b0 + p.b1*x.p.1 + p.b3*x.p.3.arr)
+  for (i in 1:n.sites){
+    for (k in 1:n.years){
+      for (j in 1:n.surveys){
+        y1[i,j,k] <- rbinom(1, size = N.tr[i,k], prob = p.tr1[i,j,k])
+  }}}
+  # fill count array with site-surv-yr x.p.3
+  p.tr2 <- plogis(p.b0 + p.b1 * x.p.1 + p.b3 * x.p.3)
+  for (i in 1:n.sites){
+    for (k in 1:n.years){
+      for (j in 1:n.surveys){
+        y2[i,j,k] <- rbinom(1, size = N.tr[i,k], prob = p.tr2[i,j,k])
+  }}}
+  # return data
+  return(list(n.sites = n.sites, n.surveys = n.surveys, n.years = n.years,
+              x.p.1 = x.p.1[,1,1], x.p.3 = x.p.3, x.p.3.mean = x.p.3.mean,
+              x.p.3.arr = x.p.3.arr, x.lam.1 = x.lam.1[,1],
+              x.lam.2 = x.lam.2[,1], yr = yr[1,], y1 = y1, y2 = y2,
+              lam.tr = lam.tr, N.tr = N.tr))
+  } #end function
R> set.seed(12345) # make results reproducible
R> sim.data <- data4sim() # create datasets
\end{verbatim}

\subsection[Real data]{Real data}
In addition to simulated data, we also demonstrate the use of \emph{R-INLA} with a real dataset in Example III. This dataset comes from a study by \cite{Kery_Royle_Schmid_2005} and is publicly available as part of the \emph{unmarked} package. The dataset includes mallard duck (\emph{Anas platyrhynchos}) counts, conducted at 239 sites on 2 or 3 occasions during the summer of 2002 as part of a Swiss program that monitors breeding bird abundance (Monitoring H\"{a}ufige Brutv\"{o}gel). In addition to counts, the dataset also includes 2 site-survey covariates related to detection (survey effort and survey date), and 3 site level covariates related to abundance (route length, elevation, and forest cover). Full dataset details are given in \cite{Kery_Royle_Schmid_2005}.

%% example 1
\section[Example I]{Example I}
\subsection[Goals]{Goals}
In Example I, we demonstrate the use of \emph{R-INLA} and compare use and performance to similar analyses using \emph{JAGS} and \emph{unmarked}. In this exercise, the functional forms of \emph{JAGS}, \emph{unmarked}, and \emph{R-INLA} models match the data generating process.  Specifically, we used the covariate x.p.3.mean to generate the count matrix y1 and analyzed the data with models that use x.p.3.mean as a covariate. This exercise was intended to demonstrate the differences and similarities in use, computation time, and estimation results across the three methods when the specified model was reasonably close to the data generating process.

\subsection[Analysis with JAGS]{Analysis with \emph{JAGS}}
We first analyzed the simulated data using \emph{JAGS}, via the \emph{runjags} package. In defining the model, we specified a Negative Binomial distribution for the abundance component, and used vague normal priors for the intercepts and the global effects of the covariates of $\lambda$ and $p$.

\begin{verbatim}
R> jags.model.string <- "
+    model {
+    # priors
+    intP ~ dnorm(0, 0.01)       # detection intercept
+    bCov1P ~ dnorm(0, 0.01)     # detection cov 1 effect
+    bCov3P ~ dnorm(0, 0.01)     # detection cov 3 effect
+    intLam ~ dnorm(0, 0.01)     # lambda intercept
+    bCov1Lam ~ dnorm(0, 0.01)   # lambda cov 1 effect
+    bCov2Lam ~ dnorm(0, 0.01)   # lambda cov 2 effect
+    bYr ~ dnorm(0, 0.01)        # year effect
+    overDisEst ~ dunif(0, 5)    # overdispersion size
+    # abundance component
+    for (k in 1:nYears){
+      for (i in 1:nSites){
+        N[i, k] ~ dnegbin(prob[i, k], overDisEst) # negative binomial
+        prob[i, k] <- overDisEst / (overDisEst + lambda[i, k])
+        log(lambda[i, k]) <- intLam + (bCov1Lam * x.lam.1[i]) + 
+                             (bCov2Lam * x.lam.2[i]) + (bYr * yr[k])
+    # detection component
+        for (j in 1:nSurveys){
+          y[i, j, k] ~ dbin(p[i,j,k], N[i,k])
+          p[i, j, k] <- exp(lp[i,j,k]) / (1 + exp(lp[i,j,k]))
+          lp[i, j, k] <- intP + (bCov1P * x.p.1[i]) + (bCov3P * x.p.3[i, k])
+        } # close j loop
+      } # close i loop
+    } # close k loop
+  } # close model loop
+  "
\end{verbatim}

Next, we defined the parameters to be monitored during the MCMC runs, bundled the data for \emph{JAGS}, and created a function for drawing random initial values for the model parameters. The initial values for abundance were made to avoid values of "NA" and zero, as these would cause computational problems.

\begin{verbatim}
R> params <- c("intP", "bCov1P", "bCov3P", "intLam", "bCov1Lam","bCov2Lam",
+              "bYr", "overDisEst") # parameters to monitor
R> jags.data <- list(y = sim.data$y1, x.lam.1 = sim.data$x.lam.1,
+               x.lam.2 = sim.data$x.lam.2, yr = sim.data$yr,
+               x.p.1 = sim.data$x.p.1, x.p.3 = sim.data$x.p.3.mean,
+               nSites = sim.data$n.sites, nSurveys = sim.data$n.surveys,
+               nYears = sim.data$n.years)
R> N.init <- sim.data$y1 # initial count values
R> N.init[is.na(N.init)] <- 1 # clean up NA's
R> N.init <- apply(N.init, c(1, 3), max) + 1 # zero values cause trouble
R> inits <- function() list(N = N.init, intLam = rnorm(1, 0, 0.01),
+               intP = rnorm(1, 0, 0.01), bCov1P = rnorm(1, 0, 0.01),
+               bCov2Lam = rnorm(1, 0, 0.01), bCov1Lam = rnorm(1, 0, 0.01),
+               bCov3P = rnorm(1, 0, 0.01), bYr = rnorm(1, 0, 0.01),
+               overDisEst = runif(1, 0.5, 2.5))
\end{verbatim}

Finally, we set the run parameters and started the MCMC process. Run parameters were chosen such that MCMC diagnostics indicated converged chains (potential scale reduction factors $<$ 1.05) and reasonably robust posterior distributions (effective sample sizes $>$ 1000). Note that the recommended number of effective samples for particularly robust inference is closer to 6000 \citep{Gong_Flegal_2016}. So MCMC processing times reported here could be considered low estimates.

\begin{verbatim}
R> nc <- 3; na <- 2500; nb <- 2500; ni <- 5000; nt <- 10 # run parameters
R> out.jags <- run.jags(model = jags.model.string, data = jags.data,
+                       monitor = params, n.chains = nc, inits = inits,
+                       burnin = nb, adapt = na, sample = ni, thin = nt,
+                       modules = "glm on", method = "parallel")
\end{verbatim}

Mean parameter estimates from the \emph{JAGS} model were reasonably close to, and not significantly different from, the input values used to generate the data (Fig. \ref{fig:fig1}). The potential scale reduction factor for all variables was $<$ 1.01, and the effective sample size for all variables was $>$ 2409. The simulation ran in parallel on 3 virtual cores, 1 MCMC chain per core, and took approximately 2104 seconds.
